---
title: "Lab 4 - Test Assignment - NYC flights"
subtitle: "TU Dortmund SoSe20 - Exploratory Data Analysis"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: haddock
    toc: no
    fig_height: 4
    fig_width: 7
margin_references: false
fontsize: 11pt
geometry: margin=.5in
lang: eng
---

## Packages

In this test assignment you will use, at a minimum, the tidyverse package. 
Load these packages on top of your document.

```{r load-packages, message=FALSE}
library(tidyverse)
library(nycflights13)
```


## Questions 

<!-- Question 1 -->

```{marginfigure}
5 points:  
- 2 pts - `count` OR `group_by` + `summarise`  
- either 1 pt - `arrange` + `desc` or `sort = TRUE` in `count` 
- 1 pt - `slice(1:10)`
- 1 pt narrative
```

1. What are the ten most common destinations for flights from NYC airports in 
2013? Your answer should be a table (or a data frame) with two columns 
(destination code and number of flights), arranged in descending order of number 
of flights. Also look up the name of the most common destination and state it 
in your narrative.

The most common destinations for flights from NYC is ___. The ten most 
common destinations are shown below.

```{r, eval = F}
___ %>%
  ___(___) %>%
  ___(___(___)) %>%
  slice(1:10)
```

<!-- Question 2 -->

```{marginfigure}
5 points:  
- 1 pts - `count` OR `group_by` + `summarise`  
- either 1 pt - `arrange` + `desc` or `sort = TRUE` in `count` 
- 2 pt - `inner_join` with airlines  
- 1 pt narrative  
```

2. What are the top three airlines with the highest numbers of flights 
departing from NYC airports in 2013? Your answer should include a table 
(or a data frame) with three columns: carrier code, number of flights, 
and name of airline, and the airlines should be listed in descending order of 
number of flights. *Hint:* The `airlines` dataset can be used to look up 
airline names from their carrier codes.

The carrier with the highest number of flights departing from NYC airports in 
2013 is XXX, followed by YYY and ZZZ.

```{r, eval = F}
___ %>%
  ___(___) %>%
  ___(___(n)) %>%
  ___(___, by = "___")
```

<!-- Question 3 -->

```{marginfigure}
10 pts:  
- 2 pt - `filter` out `NAs` (or in `median`, `na.rm = TRUE`)  
- 2 pts - output sorted in order of median aarival delay  
- 2 pts - `inner_join` with `airlines`  
- 2 pts - Frontier Airlines, 6 minutes  
- 2 pts - Alaska Airlines, -17 minutes  
```

3. Consider only flights that have non-missing arrival delay information. 
Which carrier has the highest median arrival delay and which carrier has the 
lowest median arrival delay? Your answer should include a table (or a data 
frame) with carrier code, median arrival delay, and name of airline. 
**Hint:** First calculate the median arrival delay for all carriers, then 
join with `airlines` to match airline names. For comupational efficienc, avoid 
joining the large `flights` data frame directly with `airlines`.

Highest median arrival delay is on ___ with ___ minutes, and 
lowest median arrival delay is on ___ with ___ minutes (which means 
arriving ___ minutes earlier than scheduled).

```{r, eval = F}
flights %>%
  ___(!___(___)) %>%
  ___(___) %>%
  ___(___ = median(___)) %>%
  ___(___(___)) %>%
  inner_join(___, by = "___") %>%
  slice(c(1, n())) # optional
```

<!-- Question 4 -->

```{marginfigure}
10 pts:  
- 4 pts - JFK  
- 3 pts - Jan 9  
- 3 pts - 40.835 F  
```

4. Find the flight with the highest departure delay and determine on what day 
this happened. What was the mean temperature on that day at the origin 
airport of this flight? In your answer you should include carrier code and name, 
origin airport, departure date, as well as the mean temperature on that day 
at that airport. 

First, we find the flight with the highest departure delay and take a note 
of relevant information. The highest departure delay of ___ minutes 
(approx ___ hours) was on a flight departing from ___ on ___. 

```{r, eval = F}
flights %>%
  ___(___(dep_delay)) %>%
  slice(1) %>%
  ___(___, ___, ___, ___)
```

Then, we calculate the mean temperature on that day at the origin airport. 
The average temperature on this day was ___ F. `R`

```{r, eval = F}
weather %>%
  ___(___ == ___, ___ == ___, ___ == "___") %>%
  summarise(mean_temp = ___(___))
```
