- Estimation -> Confidence interval
  - bootstrap: draw x samples from the sample with replication 
- Decision -> Hypothesis test
	- bootstrap: draw x samples from the sample with replication and measure how many below stat
  - simulate: draw x samples randomly from a theoretical distribution, e.g. binomial for testing 1 proportion
  - permute: shuffle the sample and reassign (re-assign labels) for testing two proportions

```{r}
glimpse(ncbirths)


inferred <- df %>%
  specify(response, explanatory) %>% # explanatory optional
	hypothesize(null, p, mu, med, sigma) %>% # "independence" (no options) and "point" (true prop, true mean, true median, true sd)
  generate(reps, type) %>% # type: bootstrap, simulate, or permute
  calculate(stat) 

boot_means_shifted <- ncbirths %>%
  filter(!is.na(weeks)) %>% # remove NAs
  specify(response = weeks) %>%
  hypothesize(null = "point", mu = 40) %>% # hypothesize step
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")

boot_means_shifted %>%
  filter(stat <= 38.33) %>%
  summarise(p_value = 2 * (n() / 1000))
```

inference
- estimate 
  - "mean", "median", "sum", "sd", "prop", "count", 
  - "ratio of props", "slope", and "correlation"
- test / hypothesize: indepndence or point
  - "diff in means", "diff in medians", "diff in props", 
  - "Chisq", "F", "t", "z", 

Do these data provide convincing evidence of a difference in length in gestation between mature and younger moms? Use a significance level of 10%.

```{r}
boot_diffs <- ncbirths %>%
	filter(!is.na(weeks)) %>% # remove NAs
  specify(response = weeks, explanatory = mature) %>% # explanatory optional
	hypothesize(null = "independence") %>% # "independence" (no options) and "point" (true prop, true mean, true median, true sd)
  generate(reps = 5000, type = "permute") %>% # type: bootstrap, simulate, or permute
  calculate("diff in means", order = c("mature mom", "younger mom")) # "mean", "median", "sum", "sd", "prop", "count", "diff in means", "diff in medians", "diff in props", "Chisq", "F", "t", "z", "ratio of props", "slope", and "correlation".

glimpse(boot_diffs)

ggplot(data = boot_diffs, aes(x = stat)) +
  geom_histogram(binwidth = 0.03) +
  geom_vline(xintercept = 0 - .4, color = "red") +
  geom_vline(xintercept = 0 + .4, color = "red")

boot_diffs %>%
  filter(abs(stat) >= .10 * 40) %>%
  summarise(p_value = 1 * (n() / 5000))

``` 

Estimate the difference in average lengths of gestation between mature and younger moms. Use a significance level equivalent to the hypothesis test above.

```{r}
library(tidyverse)
library(infer)

boot_diffs <- ncbirths %>%
	filter(!is.na(weeks)) %>% # remove NAs
  specify(response = weeks, explanatory = mature) %>% # explanatory optional
	# hypothesize(null = "independence") %>% # "independence" (no options) and "point" (true prop, true mean, true median, true sd)
  generate(1000, type = "bootstrap") %>% # type: bootstrap, simulate, or permute
  calculate(stat = "diff in means", order = c("mature mom", "younger mom")) # "mean", "median", "sum", "sd", "prop", "count", "diff in means", "diff in medians", "diff in props", "Chisq", "F", "t", "z", "ratio of props", "slope", and "correlation".

boot_diffs %>%
  summarize(lower_bound = quantile(stat, 0.025),
            upper_bound = quantile(stat, 0.975))
boot_diffs %>%
  summarize(lower_bound = quantile(stat, 0.05),
            upper_bound = quantile(stat, 0.95))
``` 

Do the results of the hypothesis test agree with the result of the confidence interval?

```{r}
df %>%
  specify(response, explanatory) %>% # explanatory optional
	hypothesize(null, p, mu, med, sigma) %>% # "independence" (no options) and "point" (true prop, true mean, true median, true sd)
  generate(reps, type) %>% # type: bootstrap, simulate, or permute
  calculate(stat) # "mean", "median", "sum", "sd", "prop", "count", "diff in means", "diff in medians", "diff in props", "Chisq", "F", "t", "z", "ratio of props", "slope", and "correlation".
``` 